---
title: "Model Implementation"
author: "Flannery McNair"
date: "2026-04-06"
output: html_document
---

```{r}
library(here)
library(tidyverse)
library(lubridate)
library(nlme)
library(gtsummary)
library(ggplot2)
library(car)
```


```{r}
acres_burned_2016 <- county_fires_clean %>%
  filter(YEAR_ == 2016) %>%
  aggregate(GIS_ACRES ~ NAME, FUN = sum) %>%
  rename(COUNTY = NAME)

fire_data_2016 <- inner_join(fires_2016_counts, acres_burned_2016, by = "COUNTY")

acres_burned_2018 <- county_fires_clean %>%
  filter(YEAR_ == 2018) %>%
  aggregate(GIS_ACRES ~ NAME, FUN = sum) %>%
  rename(COUNTY = NAME)

fire_data_2018 <- inner_join(fires_2018_counts, acres_burned_2018, by = "COUNTY")

acres_burned_2020 <- county_fires_clean %>%
  filter(YEAR_ == 2020) %>%
  aggregate(GIS_ACRES ~ NAME, FUN = sum) %>%
  rename(COUNTY = NAME)

fire_data_2020 <- inner_join(fires_2020_counts, acres_burned_2020, by = "COUNTY")

acres_burned_2022 <- county_fires_clean %>%
  filter(YEAR_ == 2022) %>%
  aggregate(GIS_ACRES ~ NAME, FUN = sum) %>%
  rename(COUNTY = NAME)

fire_data_2022 <- inner_join(fires_2022_counts, acres_burned_2022, by = "COUNTY")

total_2016_data <- inner_join(data_2016_SVI, fire_data_2016, by = "COUNTY")
total_2018_data <- inner_join(data_2018_SVI, fire_data_2018, by = "COUNTY")
total_2020_data <- inner_join(data_2020_SVI, fire_data_2020, by = "COUNTY")
total_2022_data <- inner_join(data_2022_SVI, fire_data_2022, by = "COUNTY")

final_dataframe <- bind_rows(
  total_2016_data %>% mutate(Year = 2016),
  total_2018_data %>% mutate(Year = 2018),
  total_2020_data %>% mutate(Year = 2020),
  total_2022_data %>% mutate(Year = 2022)
)

final_dataframe <- final_dataframe %>%
  subset(select = -c(STCNTY, E_POV150, E_HBURD, EP_POV150, EP_HBURD, EPL_POV150,
                     EPL_HBURD, EPL_UNINSUR, E_NOINT, E_AFAM, E_HISP, E_ASIAN, 
                     E_NHPI, E_TWOMORE, E_OTHERRACE, EP_NOINT, EP_AFAM, EP_HISP, 
                     EP_ASIAN, E_AIAN, EP_AIAN, EP_NHPI, EP_TWOMORE, EP_OTHERRACE))

#write_csv(final_dataframe, here("final_dataframe.csv"))
```

```{r}
df <- read.csv(here("final_dataframe.csv"), stringsAsFactors = TRUE) %>%
  subset(select = -c(E_POV, E_PCI, EP_POV, EP_PCI, EPL_POV, EPL_PCI)) %>%
  drop_na()

lm1 <- lme(CURRENT.PREVALENCE ~  GIS_ACRES + AQI, random = ~1|COUNTY, data = df)

summary(lm1)
vif(lm1)
plot(lm1)
cor.test(df$GIS_ACRES, df$CURRENT.PREVALENCE, method = "pearson")
```

random effect: county? 


```{r}
fires_per_county <- df %>%
  aggregate(n ~ COUNTY, FUN = sum)

```


Data Visualization

```{r}

```



Second, in Los Angeles County specifically, is the burned area associated with an increase in the number of days with an AQI above 100—the threshold marking conditions unhealthy for sensitive groups? We hypothesize that years with greater burned area in or near the county will show significantly more unhealthy air quality days.

need 
- daily aqi data in LA (can use AQI data) 
- fires/year, maybe monthly data?


```{r}
# finding aqi for LA 
AQI_2022_daily <- AQI_2022 %>%
  filter(`State Name` == "California") %>%
  filter(`county Name` == "Los Angeles")

AQI_2020_daily <- AQI_2020 %>%
  filter(`State Name` == "California") %>%
  filter(`county Name` == "Los Angeles")

AQI_2018_daily <- AQI_2018 %>%
  filter(`State Name` == "California") %>%
  filter(`county Name` == "Los Angeles")

AQI_2016_daily <- AQI_2016 %>%
  filter(`State Name` == "California") %>%
  filter(`county Name` == "Los Angeles")

AQI_daily <- bind_rows(
  AQI_2016_daily %>% mutate(Year = 2016),
  AQI_2018_daily %>% mutate(Year = 2018),
  AQI_2020_daily %>% mutate(Year = 2020),
  AQI_2022_daily %>% mutate(Year = 2022)
)

# days of unhealthy aqi
AQI_unhealthy <- AQI_daily %>%
  filter(Category == "Unhealthy for Sensitive Groups" | Category == "Unhealthy" |
           Category == "Very Unhealthy")

# fires in LA
LA_fires <- county_fires_clean %>%
  filter(NAME == "Los Angeles")

la_fires_plot <- ggplot(LA_fires, aes(x =  factor(year(ALARM_DATE)))) + 
  geom_bar(stat = "count")

la_fires_plot

AQI_plot <- ggplot(AQI_unhealthy, aes(x = factor(year(Date)), y = AQI)) + 
  geom_boxplot()

AQI_plot

# number of each type of aqi each year
df1 <-AQI_daily %>% 
  count(year(Date), Category)

ggplot(df1, aes(x=factor(`year(Date)`), y=n, fill=Category))+  
  geom_bar(stat="identity")
```
consolidate into months? 
if i have the length of the fire, can i count the number of days that fires are present in each county? and on specific days? 
maybe do months first 

```{r}
monthly_fire_data <- LA_fires %>% 
  mutate(month_year = format(ALARM_DATE, "%Y-%m")) %>%
  aggregate(GIS_ACRES ~ month_year, FUN = sum)

monthly_AQI_data <- AQI_daily %>%
  mutate(month_year = format(Date, "%Y-%m")) %>%
  aggregate(AQI ~ month_year, FUN = mean)

monthly_data <- left_join(monthly_AQI_data, monthly_fire_data, by = "month_year") %>%
  mutate(across(where(is.numeric), ~replace_na(., 0)))

aqi_vs_fire <- ggplot(monthly_data, aes(x = month_year, y = GIS_ACRES)) + 
  geom_bar(stat = "identity", fill = "red") + 
  geom_point(data = monthly_data, aes(x = month_year, y = AQI*100), color = "blue") + 
  scale_y_continuous(name = "Fire Counts", 
                     sec.axis = sec_axis(~ . / 100, name = "AQI")) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

aqi_vs_fire
  
lm3 <- lm(AQI ~ GIS_ACRES, data = monthly_data)
summary(lm3)
plot(lm3)
```

