---
title: "Model Implementation"
author: "Flannery McNair"
date: "2026-04-06"
output: html_document
---

```{r}
library(here)
library(tidyverse)
library(lubridate)
library(nlme)
library(gtsummary)
library(ggplot2)
library(car)
library(DHARMa)
```


```{r}
acres_burned_2016 <- county_fires_clean %>%
  filter(YEAR_ == 2016) %>%
  aggregate(GIS_ACRES ~ NAME, FUN = sum) %>%
  rename(COUNTY = NAME)

fire_data_2016 <- inner_join(fires_2016_counts, acres_burned_2016, by = "COUNTY")

acres_burned_2018 <- county_fires_clean %>%
  filter(YEAR_ == 2018) %>%
  aggregate(GIS_ACRES ~ NAME, FUN = sum) %>%
  rename(COUNTY = NAME)

fire_data_2018 <- inner_join(fires_2018_counts, acres_burned_2018, by = "COUNTY")

acres_burned_2020 <- county_fires_clean %>%
  filter(YEAR_ == 2020) %>%
  aggregate(GIS_ACRES ~ NAME, FUN = sum) %>%
  rename(COUNTY = NAME)

fire_data_2020 <- inner_join(fires_2020_counts, acres_burned_2020, by = "COUNTY")

acres_burned_2022 <- county_fires_clean %>%
  filter(YEAR_ == 2022) %>%
  aggregate(GIS_ACRES ~ NAME, FUN = sum) %>%
  rename(COUNTY = NAME)

fire_data_2022 <- inner_join(fires_2022_counts, acres_burned_2022, by = "COUNTY")

total_2016_data <- inner_join(data_2016_SVI, fire_data_2016, by = "COUNTY")
total_2018_data <- inner_join(data_2018_SVI, fire_data_2018, by = "COUNTY")
total_2020_data <- inner_join(data_2020_SVI, fire_data_2020, by = "COUNTY")
total_2022_data <- inner_join(data_2022_SVI, fire_data_2022, by = "COUNTY")

final_dataframe <- bind_rows(
  total_2016_data %>% mutate(Year = 2016),
  total_2018_data %>% mutate(Year = 2018),
  total_2020_data %>% mutate(Year = 2020),
  total_2022_data %>% mutate(Year = 2022)
)

final_dataframe <- final_dataframe %>%
  subset(select = -c(STCNTY, E_POV150, E_HBURD, EP_POV150, EP_HBURD, EPL_POV150,
                     EPL_HBURD, EPL_UNINSUR, E_NOINT, E_AFAM, E_HISP, E_ASIAN, 
                     E_NHPI, E_TWOMORE, E_OTHERRACE, EP_NOINT, EP_AFAM, EP_HISP, 
                     EP_ASIAN, E_AIAN, EP_AIAN, EP_NHPI, EP_TWOMORE, EP_OTHERRACE))

all_data <- final_dataframe %>%
  subset(select = c(COUNTY, AQI, YEARS, `CURRENT PREVALENCE`, RPL_THEME1, RPL_THEME2,
                    RPL_THEME3, RPL_THEME4, RPL_THEMES, n, GIS_ACRES)) %>%
  rename(Fire_Count = n)

write_csv(all_data, here("LiangCMcNairFPereiraA_dataframe.csv"))
```

```{r}
df <- read.csv(here("final_dataframe.csv"), stringsAsFactors = TRUE) %>%
  subset(select = -c(E_POV, E_PCI, EP_POV, EP_PCI, EPL_POV, EPL_PCI)) %>%
  drop_na() %>%
  mutate(
    # Log10 transform burned area; zero-fire county-years set to log10(1) = 0
    GIS_ACRES     = ifelse(is.na(GIS_ACRES) | GIS_ACRES == 0, 1, GIS_ACRES),
    log_acres = log10(GIS_ACRES)
  )

lm1 <- lme(CURRENT.PREVALENCE ~  GIS_ACRES + AQI, random = ~1|COUNTY, data = df)

summary(lm1)
vif(lm1)
plot(lm1)
cor.test(df$GIS_ACRES, df$CURRENT.PREVALENCE, method = "pearson")
```

random effect: county? 


```{r}
fires_per_county <- df %>%
  aggregate(n ~ COUNTY, FUN = sum)

```


Data Visualization

```{r}
burned_area_before <- ggplot(df, aes(x = GIS_ACRES)) +
  geom_histogram(bins = 40, fill = "orange", color = "white", linewidth = 0.2) +
  labs(title = "Acres Burned", x = "Acres", y = "Count") +
  theme_bw()
burned_area_before

asthma_graph <- ggplot(df, aes(x = CURRENT.PREVALENCE)) +
  geom_histogram(fill = "salmon", color = "white", linewidth = 0.2) +
  labs(title = "Asthma Prevalence", x = "Asthma Prevalence (%)", y = "Count") +
  theme_bw()
asthma_graph

AQI_graph <- ggplot(df, aes(x = AQI)) + 
  geom_histogram()
AQI_graph

SVI_graph <- ggplot(df, aes(x = RPL_THEMES)) + 
  geom_histogram(fill = "darkgreen", color = "white", linewidth = 0.2) +
  labs(title = "SVI", x = "RPL_THEME", y = "Count") +
  theme_bw()
SVI_graph

```



Second, in Los Angeles County specifically, is the burned area associated with an increase in the number of days with an AQI above 100—the threshold marking conditions unhealthy for sensitive groups? We hypothesize that years with greater burned area in or near the county will show significantly more unhealthy air quality days.

need 
- daily aqi data in LA (can use AQI data) 
- fires/year, maybe monthly data?


```{r}
# finding aqi for LA 
AQI_2022_daily <- AQI_2022 %>%
  filter(`State Name` == "California") %>%
  filter(`county Name` == "Los Angeles")

AQI_2020_daily <- AQI_2020 %>%
  filter(`State Name` == "California") %>%
  filter(`county Name` == "Los Angeles")

AQI_2018_daily <- AQI_2018 %>%
  filter(`State Name` == "California") %>%
  filter(`county Name` == "Los Angeles")

AQI_2016_daily <- AQI_2016 %>%
  filter(`State Name` == "California") %>%
  filter(`county Name` == "Los Angeles")

AQI_daily <- bind_rows(
  AQI_2016_daily %>% mutate(Year = 2016),
  AQI_2018_daily %>% mutate(Year = 2018),
  AQI_2020_daily %>% mutate(Year = 2020),
  AQI_2022_daily %>% mutate(Year = 2022)
)

# days of unhealthy aqi
AQI_unhealthy <- AQI_daily %>%
  filter(Category == "Unhealthy for Sensitive Groups" | Category == "Unhealthy" |
           Category == "Very Unhealthy")

# fires in LA
LA_fires <- county_fires_clean %>%
  filter(NAME == "Los Angeles")

la_fires_plot <- ggplot(LA_fires, aes(x =  factor(year(ALARM_DATE)))) + 
  geom_bar(stat = "count")

la_fires_plot

AQI_plot <- ggplot(AQI_unhealthy, aes(x = factor(year(Date)), y = AQI)) + 
  geom_boxplot()

AQI_plot

# number of each type of aqi each year
df1 <-AQI_daily %>% 
  count(year(Date), Category)

ggplot(df1, aes(x=factor(`year(Date)`), y=n, fill=Category))+  
  geom_bar(stat="identity") + 
  labs(title = "Air Quality in LA County", x = "Year", 
       y = "Count") + 
  theme_bw()
```

```{r}
#lm_a <- lme(AQI_daily$Category ~ )
```

consolidate into months? 
if i have the length of the fire, can i count the number of days that fires are present in each county? and on specific days? 
maybe do months first 

```{r}
monthly_fire_data <- LA_fires %>% 
  mutate(month_year = format(ALARM_DATE, "%Y-%m")) %>%
  aggregate(GIS_ACRES ~ month_year, FUN = sum) %>%
  mutate(
    # Log10 transform burned area; zero-fire county-years set to log10(1) = 0
    GIS_ACRES     = ifelse(is.na(GIS_ACRES) | GIS_ACRES == 0, 1, GIS_ACRES),
    log_acres = log10(GIS_ACRES)
  )
  

monthly_AQI_data <- AQI_daily %>%
  mutate(month_year = format(Date, "%Y-%m")) %>%
  group_by(month_year) %>%
  summarise(
    mean_AQI = mean(AQI, na.rm = TRUE),
    days_above_100 = sum(AQI > 100, na.rm = TRUE))

monthly_data <- left_join(monthly_AQI_data, monthly_fire_data, by = "month_year") %>%
  mutate(across(where(is.numeric), ~replace_na(., 0)))  %>%
  mutate(date_column = ym(month_year))

write_csv(monthly_data, here("LiangCMcNairFPereiraA_dataframe_monthly.csv"))

aqi_vs_fire <- ggplot(monthly_data, aes(x = month_year)) + 
  geom_bar(aes(y = GIS_ACRES, fill = "Acres Burned"), stat = "identity") + 
  geom_point(aes(y = days_above_100 * 1000, color = "Days AQI > 100")) + 
  scale_y_continuous(
    name = "Total Acres Burned", 
    sec.axis = sec_axis(~ . / 1000, name = "# of Days with AQI > 100")
  ) + 
  scale_fill_manual(values = c("Acres Burned" = "red")) +
  scale_color_manual(values = c("Days AQI > 100" = "blue")) +
  scale_x_discrete(breaks = monthly_data$month_year[seq(1, nrow(monthly_data), by = 3)]) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.position = "bottom") + 
  labs(
    x = "Month-Year", 
    title = "Relationship between Total Acres Burned and Number of Days 
with an AQI Over 100",
    subtitle = "2016, 2018, 2020, 2022",
    fill = "",
    color = ""
  )

#    scale_x_date(date_breaks = "3 months", date_labels = "%Y-%m") +

aqi_vs_fire
  
lm3 <- glm(days_above_100 ~ log_acres, data = monthly_data, family = "poisson")
summary(lm3)
plot(lm3)
cor.test(monthly_data$days_above_100, monthly_data$log_acres, method = "pearson")

aqi_vs_fire_dot <- ggplot(monthly_data, aes(log_acres, days_above_100)) +
  geom_point() +
  geom_smooth(method = "lm") + 
  stat_cor(aes(label = after_stat(rr.label))) + 
  labs(y = "Days Above 100 AQI", x = "log10(Acres Burned)",
       title = "Log10 Acres Burnes vs Days Above 100 AQI") + 
  theme_bw()

aqi_vs_fire_dot

tbl_regression(lm3)
```


lme(mean_aqi ~ log_acres + svi + year, random = ~1 | county,
    data   = cleaned_data_rq1,method = "ML")
